### 2018-07-09 latex/figurelegends.tex
Minor formatting changes before sending manuscript for internal review
revised introduction and figure legends, still working on results

\begin{figure}[ht]
  \centering
  \includegraphics{fig_1_pathway_overview}
  \caption{\textbf{Schematic overview of NAD biosynthesis pathways} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR). Nam is the main precursor in human and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases). For the recycling of Nam, two different pathways exist. The pathway found in yeast and many bacteria starts with the deamination of Nam by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT). The Nam N-methyltransferase (NNMT) degrades Nam to methyl-Nam (MNam), which is in mammals excreted with the urine.}
  \label{fig:pathway_overview}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected major taxa. NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa. Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour, which is explained in the lower right of the figure. Below the taxon name, the number of species in that taxon is given. B) Common tree of selected taxa within the Metazoa, including 334 species. The pie charts indicate the distribution of species within the respective taxon that encode the enzyme combination indicated by the different colours. The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon. The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa. The branch length is arbitrary.}
  \label{fig:phylo_distribution}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see experimental procedures) to simulate NAD consumption flux. The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature. All other enzyme concentrations were set equal. Details are given in supplementary table~S2. In the absence of NNMT (blue lines), steady state NAD consumption rates (A) are higher despite reduced NAD concentrations (B). Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
  \label{fig:NNMT_NAD_flux}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates. All the parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}. In the absence of NNMT (A and B), the $K_{M}$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates. In the presence of NNMT (C and D), decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux and NAD concentration. NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates. Comparing the situation with and without NNMT (E and F) at different NamPRT $K_{M}$ reveals that at low $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
  \label{fig:NamPRT_affinity_Nam}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA. For a more comprehensive alignment, please see supplementary figure~S1. The visualisation of human NamPRT (B) is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}. The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. Immunofluorescence images show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop (C) is not changed compared to FLAG-tagged human wildtype NamPRT. Both show a heterogeneous nuclear cytosolic localisation in HeLa S3 cells. \textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E). Bars in D) and E) with different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
  \label{fig:unresolved_loop}
\end{figure}

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics{fig_1_pathway_overview}
  \caption{\textbf{Schematic overview of NAD biosynthesis pathways} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR). Nam is the main precursor in human and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases). For the recycling of Nam, two different pathways exist. The pathway found in yeast and many bacteria starts with the deamidation of Nam by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT). The Nam N-methyltransferase (NNMT) degrades Nam to methyl-Nam (MNam), which is in mammals excreted with the urine.}
  \label{fig:pathway_overview}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected major taxa. NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa. Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure. Below the taxon name, the number of species in that taxon is given. B) Common tree of selected taxa within the Metazoa, including 334 species. The pie charts indicate the distribution of species within the respective taxon that encode the enzyme combination indicated by the different colours. The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon. The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa. The branch length is arbitrary.}
  \label{fig:phylo_distribution}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see experimental procedures) to simulate steady state NAD  consumption flux  (A) and concentration (B). The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature. All other enzyme concentrations were set equal. Details are given in supplementary table~S2. In the absence of NNMT (blue lines), steady state NAD consumption rates are higher despite reduced NAD concentrations. Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
  \label{fig:NNMT_NAD_flux}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.72\textwidth]{fig_4_NamPRT_affinity_Nam}
  \caption{\textbf{Role of NamPRT substrate affinity} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates. All parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}. In the absence of NNMT, the $K_{M}$ of NamPRT has little influence on NAD consumption (A) and concentration (B), but both are changing with cell division rates. In the presence of NNMT, decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux (C) and NAD concentration (D). NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates. Comparing the situation with and without NNMT (E and F) at different NamPRT $K_{M}$ reveals that at low $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
  \label{fig:NamPRT_affinity_Nam}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA. For a more comprehensive alignment, please see supplementary figure~S1. The visualisation of human NamPRT (B) is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}. The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. Immunofluorescence images (C) show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop  is not changed compared to FLAG-tagged human wildtype NamPRT. Both show a heterogeneous nuclear-cytosolic localisation in HeLa S3 cells. \textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E). Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
  \label{fig:unresolved_loop}
\end{figure}

### 2018-07-09 latex/introduction.tex
revised introduction and figure legends, still working on results

The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated. However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD homeostasis have been linked to various diseases, in particular related to ageing, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}. Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. In the first step of the salvage pathway, Nam is converted to the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This extraordinary affinity appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \todo{Needs reference}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).

We here present a phylogenetic analysis of the NAD pathways. The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost. This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam. Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.

>>>>

The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated. It has now been established, however, that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018}. This  suggests that NAD-dependent signalling reactions consume substantial amounts of NAD. Therefore, we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD homeostasis have been linked to various mainly age related diseases, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}. Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway staring with the convertion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{PMID:19666527}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. Although it has been suggested that it potentially acts as methylation sink \cite{NNMTreview2018} . Methyl-Nam produced by NNMT is in mammals excreted with the urine, thus removing Nam from recycling.

Secondly, in many prokayotes as well as in plants and fungi a four steps starting with the deamidation of Nam to nicotinic acid (NA) by NADA is used. (\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT. The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{Gossmann2011}. The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This reaction includes an enzyme adenylation step that consumes ATP. Therefore the Nam recycling by NADA is energetically less efficient under normal conditions.

We here present a phylogenetic analysis of the NAD pathways. Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost. This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam. Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT. Simulating the resource competition, we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.

### 2018-07-09 latex/results.tex
revised introduction and figure legends, still working on results

\subsection{Paradoxical evolutionary correlation between NAD-dependent signaling and precursor metabolism}

Two principal pathways exist that recycle Nam. As shown in figure \ref{fig:pathway_overview}, the major pathway found in most prokaryotes, yeast, and plants includes four steps starting with the deamidation of Nam to nicotinic acid by NADA. The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into Nam mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme, NAPRT, converts NA into the corresponding mononucleotide (NAMN) in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This reaction includes an enzyme adenylation step that consumes ATP. Even though Nam recycling by NADA is energetically less efficient, this pathway is preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).

Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail revealed that not only NamPRT replaced NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were unable to find any \textit{NNMT} gene with considerable \todo{define considerable} similarity in fungi or plants. Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the \textit{NNMT} of nematodes or deuterostomes. The yeast NNMT protein has meanwhile been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.

>>>>

\subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
As the Preiss-Handler pathway  is  present in virtually all organisms analysed previously\cite{Gossmann2012FEBS} and as based on metabolic pathway analysis we have previously shown tha NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor use \cite{Gossmann2011} we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways NADA and NamPRT as well as on the NAM methyltransferase NNMT as the distribution of these enzymes seems to differ substantially between different organisms. As can be seen from Fig. (fig.~\ref{fig:phylo_distribution}A, bacteria, fungi and plants predominantly encode NADA and only a few of them harbour NamPRT.  In contrast, Metazoa predominantly have NamPRT together with NNMT. Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT replaced by NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were unable to find any \textit{NNMT} gene with considerable \todo{define considerable} similarity in fungi or plants.
